---
title: "Data Science Tutorial: Intro to Animations in R"
author: "Nolan & Anton"
date: "9/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Animations can be very powerful tools when used effectively, allowing you to see progression over time, iterate through plots, and help you tell a better story with the data. This guide will give an overview of packages, techniques, and tips for creating animated plots in R. 

In this guide you will learn:

- Necessary packages for creating animated plots
- How to turn static plots into animations
- What each argument in the functions do
- Ways to edit, improve, and save your animations



# Getting Started

## Packages

To create animations, there are some R packages you will need to install & load first. The main packages we will be using are:

1. gganimate
2. ggplot2
3. dplyr
4. plotly
5. gapminder

You will likely need to install the following packages if you haven't worked with animations before:

```{r, eval=F}
install.packages("gganimate")
install.packages("plotly")
install.packages("gapminder")
```

Then load the necessary packages:

```{r}
library(ggplot2)
library(dplyr)
library(gganimate)
library(plotly)
library(gapminder)
```


## Data

To create animated plots we must first have some tidy data to work with. We'll start with an example data set to get familiar with some of the features before trying it on unique data.

A great example data set for animations is gapminder, which contains data on worldwide economic, health, and other public information. Load the data set and preview it below:

```{r}
head(gapminder)
```

Before we can animate a plot, we must first make a regular one in ggplot. Create a plot using some of the 6 variables in the data set.

Here's an example:
```{r}
gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  labs(x = "GDP per capita", y = "Life expectancy") +
  theme_classic()
```

Now we're ready for animations.


# Creating Animations

