---
title: "Data Science Tutorial: Intro to Animations in R"
author: "Nolan & Anton"
date: "9/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Animations can be very powerful tools when used effectively, allowing you to see progression over time, iterate through plots, and help you tell a better story with the data. This guide will give an overview of packages, techniques, and tips for creating animated plots in R. 

In this guide you will learn:

- Necessary packages for creating animated plots
- How to turn static plots into animations
- Components of gganimate and what each function does
- Ways to edit, improve, and save your animations



# Getting Started

## Packages

To create animations, there are some R packages you will need to install & load first. The main packages we will be using are:

1. gganimate
2. ggplot2
3. tidyverse
4. gapminder

You will likely need to install the following packages if you haven't worked with animations before:

```{r, eval=F}
install.packages("gganimate")
install.packages("gapminder")
```

Then load the necessary packages:

```{r}
library(gganimate)
library(ggplot2)
library(tidyverse)
library(gapminder)
```


## Data

To create animated plots we must first have some tidy data to work with. We'll start with an example data set to get familiar with some of the features before trying it on unique data.

A great example data set for animations is gapminder, which contains data on worldwide economic, health, and other public information. Load the data set and preview it below:

```{r}
head(gapminder)
```

Before we can animate a plot, we must first make a regular one in ggplot. Create a plot using some of the 6 variables in the data set.

Here's an example:

```{r}
gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  labs(x = "GDP per capita", y = "Life expectancy") +
  theme_classic()
```

Now we're ready for animations.


# Creating Animations

Within gganimate there are 6 key components that can be altered within an animation.

- Transitions & transition states
- Views
- Shadows
- Easing
- Enter and Exit
- Rendering

We will walk through each of these in detail and build upon them to create some cool visualizations.



## Transitions & transition states

### Basics

Transitions control how the data gets displayed in your animation. Do you want your data to populate over time by year, by category, or another filter? This is determined by the transition function you select. The list of transition functions is shown below, along with their functionality.

| Name        | Functionality |
|:-----------|:-----------|
| transition_components | Transition individual components through their own lifecycle |
| transition_events | Transition individual events in and out |
| transition_filter | Transition between different filters |
| transition_layers | Build up a plot, layer by layer |
| transition_manual | Create an animation by specifying the frame membership directly |
| transition_reveal| Reveal data along a given dimension | 
| transition_states | Transition between several distinct stages of the data |
| transition_time | Transition through distinct states in time |


Transition functions are simply added to your regular ggplot statements following a "+". Lets try using transition_time() first. The transition length will correspond to the actual time difference between the data. This can be a good option when you have yearly data. Add **transition_time()** to your plot, with the year variable as the argument.

```{r, eval=F}
gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  labs(x = "GDP per capita", y = "Life expectancy") +
  theme_classic() +
  transition_time(year)
```

You should now see your plot animated in the viewer in the bottom right panel of your screen!

While the points are moving, it may be hard to follow them as they are close together and all look the same. Add some color and/or size to your plot so we can see the animation better.

```{r, eval=F}
gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, color = country, size = pop)) +
  geom_point(alpha = 0.75, show.legend = FALSE) +
  scale_color_viridis_d() +
  labs(x = "GDP per capita", y = "Life expectancy") +
  theme_classic() +
  transition_time(year)
```


### Try it out

Head over to (Google Search Trends)[https://trends.google.com/trends/?geo=US] and search up a term that interests you. You should get a graph of the popularity of the search term over time. Click "add comparison" at the top and add at least two more terms (so you have 3 or more total).

One you have a graph of multiple terms, download this as a .csv file to your computer.

**Note:** Once you have your .csv downloaded, click on it in the bottom right "Files" pane and click "View File". The .csv file will open in another tab. Delete the first two rows so that "Week" and the terms you searched are now at the top. Click "Save", and now your .csv is ready for importing.

```{r}
popularity <- read_csv("/Users/nolan/MACALESTER/Stat 456/multiTimeline.csv")
head(popularity)
```

We want this data in tidy format, so lets **pivot_longer** our data.

```{r}
popularity <- popularity %>% 
  pivot_longer(cols = `World cup`:`UEFA Champions League`,
               names_to = "Search Term",
               values_to = "Popularity")
```

Now plot the popularity over time:

```{r}
popularity %>% 
  ggplot(aes(x = Week, y = Popularity, group = `Search Term`, color = `Search Term`)) +
  geom_line() +
  labs(x = "Time", y = "Popularity") +
  theme_classic() +
  theme(legend.position = "top")
```

From here, try out some other transition functions like **transition_reveal()** or others.

```{r, eval = F}
popularity %>% 
  ggplot(aes(x = Week, y = Popularity, group = `Search Term`, color = `Search Term`)) +
  geom_line() +
  labs(x = "Time", y = "Popularity") +
  theme_classic() +
  theme(legend.position = "top") +
  transition_reveal(Week)
```



### Labels

Another thing we should make sure to add to our animation is how the animation is transitioning. While we know the animation is changing by year, our viewer may not realize or understand that. With some transition functions this may or may not be easily  inferable, so it's good practice to include this information on the plot. We can add this information in the **labs()** part of our ggplot statement, specifying one of the corresponding transition variables related to the transition function we are using (see below).

|Name |	Label variable(s) |
|:---------|:---------|
| transition_components |	frame_time |
| transition_events	| frame_time |
| transition_filter	| previous_filter, closest_filter, next_filter |
| transition_layer	| previous_layer, closest_layer, next_layer, nlayers |
| transition_manual	| previous_frame, current_frame, next_frame |
| transition_reveal	| frame_along |
| transition_states	| previous_state, closest_state, next_state |
| transition_time	| frame_time |

In this case, lets add **title = "Year: {frame_time}"** to our label.

```{r, eval=F}
gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, color = country, size = pop)) +
  geom_point(alpha = 0.75, show.legend = FALSE) +
  scale_color_viridis_d() +
  labs(x = "GDP per capita", y = "Life expectancy", title = "Year: {frame_time}") +
  theme_classic() +
  transition_time(year)
```
