---
title: "Data Science Tutorial: Intro to Animations in R"
author: "Nolan & Anton"
date: "9/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Animations can be very powerful tools when used effectively, allowing you to see progression over time, iterate through plots, and help you tell a better story with the data. This guide will give an overview of packages, techniques, and tips for creating animated plots in R. 

In this guide you will learn:

- Necessary packages for creating animated plots
- How to turn static plots into animations
- What each argument in the functions do
- Ways to edit, improve, and save your animations



## Packages

To create animations, there are some R packages you will need to install & load first. The main packages we will be using are:

1. gganimate
2. ggplot2
3. plotly

You will likely need to install the following packages if you haven't worked with animations before:

```{r, eval=F}
install.packages("gganimate")
install.packages("plotly")
```

```{r}
library(ggplot2)
library(gganimate)
library(plotly)
```


