---
title: "Data Science Tutorial: Intro to Animations in R"
author: "Nolan & Anton"
date: "9/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Animations can be very powerful tools when used effectively, allowing you to see progression over time, iterate through plots, and help you tell a better story with the data. This guide will give an overview of packages, techniques, and tips for creating animated plots in R. 

In this guide you will learn:

- Necessary packages for creating animated plots
- How to turn static plots into animations
- Components of gganimate and what each function does
- Ways to edit, improve, and save your animations



# Getting Started

## Packages

To create animations, there are some R packages you will need to install & load first. The main packages we will be using are:

1. gganimate
2. ggplot2
3. dplyr
4. plotly
5. gapminder

You will likely need to install the following packages if you haven't worked with animations before:

```{r, eval=F}
install.packages("gganimate")
install.packages("plotly")
install.packages("gapminder")
```

Then load the necessary packages:

```{r}
library(ggplot2)
library(dplyr)
library(gganimate)
library(plotly)
library(gapminder)
```


## Data

To create animated plots we must first have some tidy data to work with. We'll start with an example data set to get familiar with some of the features before trying it on unique data.

A great example data set for animations is gapminder, which contains data on worldwide economic, health, and other public information. Load the data set and preview it below:

```{r}
head(gapminder)
```

Before we can animate a plot, we must first make a regular one in ggplot. Create a plot using some of the 6 variables in the data set.

Here's an example:

```{r}
gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  labs(x = "GDP per capita", y = "Life expectancy") +
  theme_classic()
```

Now we're ready for animations.


# Creating Animations

Within gganimate there are 6 key components that can be altered within an animation.

- Transitions & transition states
- Views
- Shadows
- Easing
- Enter and Exit
- Rendering

We will walk through each of these in detail and build upon them to create some cool visualizations.



## Transitions & transition states

Transitions control how the data gets displayed in your animation. Do you want your data to populate over time by year, by category, or another filter? This is determined by the transition function you select. The list of transition functions is shown below, along with their functionality.

| Name        | Functionality |
|:-----------|:-----------|
| transition_components | Transition individual components through their own lifecycle |
| transition_events | Transition individual events in and out |
| transition_filter | Transition between different filters |
| transition_layers | Build up a plot, layer by layer |
| transition_manual | Create an animation by specifying the frame membership directly |
| transition_null | Keep all data constant across the animation |
| transition_reveal| Reveal data along a given dimension | 
| transition_states | Transition between several distinct stages of the data |
| transition_time | Transition through distinct states in time |


