---
title: "FP4"
author: "Nolan & Anton"
date: "11/8/2022"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
```


# Deliverables

- 6-8 min presentation
- Analysis of research questions & findings
- Project plan for rest of semester


# Research Questions

1.

  - Focusing on federal aid & loans, are there any gender and race biases in how much aid is given based on EFC?

  - At what point does your family income (and/or EFC) result in zero federal aid?
  
2.

  - What majors provide the best ROI of expected family contributions in terms of salary post grad?
  
  - What majors provide the biggest increase in financial well-being based on family income or EFC?




# Data

```{r}
data <- read_csv("11_8_data.csv")
head(data)
```

```{r}
data <- readxl::read_xlsx("11_8_data.xlsx", sheet = "salary")
conf_int <- readxl::read_xlsx("11_8_data.xlsx", sheet = "confidence_intervals")

data %>% 
  ggplot() +
  geom_bar(aes(x = fct_reorder(`Field of study: undergraduate (23 categories)`, `Primary job: Annualized salary, 2012`, .desc = T), y = `Primary job: Annualized salary, 2012`), stat = "identity", fill = "skyblue", alpha = 0.7) +
  geom_errorbar(aes(x = `Field of study: undergraduate (23 categories)`, ymin=`CI_low`, ymax=`CI_high`), width=0.4, colour="orange", alpha=0.9, size=1) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 52, vjust = 1, hjust=1, size = 5.5, color = "black")) +
  labs(x = "Field of Study, 2008", y = "Salary, 2012")
```





