---
title: "Domain Expert Meeting"
author: "Nolan & Anton"
date: "10/28/2022"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r}
library(readxl)
library(ggplot2)
library(tidyverse)
```


# Research Questions:

1. Who benefits the most from the loan forgiveness (gender, race, class)?

2. How has the cost of tuition tracked in comparison to inflation rates, cost of living, and post-college income?

3. What majors provide most financial security and “easiest” opportunities to pay off debt? Which majors require students to take out most money relative to cost of attendance?



# Preliminary Results & Analysis:


## How has the cost of college changed over time?

- **Data source**: National Center for Education Statistics, [https://nces.ed.gov/programs/digest/d21/tables/dt21_330.10.asp?current=yes]()
- **Data description**: This data set includes the "average undergraduate tuition, fees, room, and board rates charged for full-time students in degree-granting post secondary institutions, by level and control of institution: Selected years, 1963-64 through 2020-21".

```{r}
tuition_cost <- read_excel("college_cost_over_time.xls", sheet = "Sheet1")
#head(tuition_cost)

tuition_cost %>% 
  ggplot(aes(x = Year, y = Amount, color = Length)) +
  geom_line() +
  facet_wrap(~Type) +
  theme_classic() +
  labs(title = "Average undergraduate cost of fees, room, and board rates in current USD", x = "", y = "")
```



## Who borrows the most money for college?

- **Data source**: National Center for Education Statistics [https://nces.ed.gov/datalab/powerstats/121-national-postsecondary-student-aid-study-2016-undergraduates/averages-medians-percents]()
- ***Data description**: - this data table includes the "cumulative amount borrowed for undergrad (average) by Race/ethnicity and gender". This data was collected in the 2016 addition of the National Postsecondary Student Aid Study.

```{r}
Cum_amt_borrowed_2016 <- read_excel("amt_borrowed.xlsx", sheet = "Sheet1")
#head(Cum_amt_borrowed_2016)

Cum_amt_borrowed_2016 %>% 
  ggplot(aes(x = fct_reorder(Race, Amount, .desc = T), y = Amount, fill = Gender)) +
  geom_bar(stat="identity", position = position_dodge()) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust=1, size = 8.5, color = "black")) +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "", y = "", fill = "", title = "Average amount borrowed for undergrad in 2016 (USD) by race and gender")
```


## Who recieves the most aid in paying for college?

```{r}
total_aid_2016 <- read_excel("total_aid.xlsx")
#head(total_aid_2016)

total_aid_2016 %>% 
  ggplot(aes(x = fct_reorder(`Race/ethnicity`, `Total aid amount`, .desc = T), y = `Total aid amount`, fill = Gender)) +
  geom_bar(stat="identity", position = position_dodge()) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust=1, size = 8.5, color = "black")) +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "", y = "", fill = "", title = "Average total aid amount for undergrad in 2016 (USD) by race and gender")
```


## How does the amount borrowed for college vary by major and parent's income?

```{r}
#devtools::install_github("slowkow/ggrepel")
library(ggrepel)
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
```

```{r}
vizdata <- read_csv("viz1data.csv")
```

```{r}
vizdata1 <- vizdata %>%
  filter(field !="Total")
```

```{r}
set.seed(123)
ggplot(vizdata1, aes(x=mean_borrow,y=mean_income,label=field, color=major_category)) +
  geom_point(size = 2.5) +
  geom_text_repel(aes(label=field), 
                  size=2.25, 
                  box.padding = 0.5, 
                  segment.size = 0.25,
                  color = "black") +
  xlim(3500,15000) + 
  ylim(60000,110000) +
  theme_classic() +
  labs(x = "Mean Cumulative Amount Borrowed (USD)", 
       y = "Mean Parents' Income (USD)", 
       title = "Amount Borrowed for Undergrad Majors Compared to Parents' Income, 2016") +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "hide")
```



## What majors tend to lead to higher paying jobs after college?

```{r}
salary_by_major_2018 <- read_excel("salary_by_major_0818.xlsx")
#head(salary_by_major_2018)

salary_by_major_2018 %>% 
  filter(`Field of study: undergraduate (10 categories)` != "Undeclared") %>% 
  ggplot(aes(x = fct_reorder(`Field of study: undergraduate (10 categories)`, `Current job, as of B&B:08/18 interview: Annualized salary`, .desc = F), y = `Current job, as of B&B:08/18 interview: Annualized salary`)) +
  geom_bar(stat="identity", fill = "#69b3a2") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1, size = 7, color = "black")) +
  labs(x = "Undergraduate Field of Study, 2008", y = "Current Job Salary in 2018", title = "")
```
