---
title: "Domain Expert Meeting"
author: "Nolan & Anton"
date: "10/28/2022"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r}
library(readxl)
library(ggplot2)
library(tidyverse)
```


# Research Questions

1. With rising tuition costs, who is college most affordable for, and what fields of study do they pursue?

2. Which college majors are worthwhile to study from a financial standpoint? (Which majors provide the most financial security & “easiest” opportunities to pay off debt?)

3. Which college majors provide the biggest uplift in a student's financial situation from pre-college to post-graduation? (Biggest change in financial indicators, i.e. salary)



# Preliminary Results & Analysis


### How has the cost of college changed over time?

- **Data source**: National Center for Education Statistics, [https://nces.ed.gov/programs/digest/d21/tables/dt21_330.10.asp?current=yes]()
- **Data description**: This data set includes the "average undergraduate tuition, fees, room, and board rates charged for full-time students in degree-granting post secondary institutions, by level and control of institution: Selected years, 1963-64 through 2020-21".

```{r}
tuition_cost <- read_excel("college_cost_over_time.xls", sheet = "Sheet1")
#head(tuition_cost)

tuition_cost %>% 
  ggplot(aes(x = Year, y = Amount, color = Length)) +
  geom_line() +
  facet_wrap(~Type) +
  theme_classic() +
  labs(title = "Average undergraduate cost of fees, room, and board rates in current USD", x = "", y = "")
```

- **Findings**: The cost of college has increased rapidly, especially at private institutions. At face value, this increase makes college less affordable for many students, and may mean they have to take out more loans and debt to finance it. 


### Who borrows the most money for college?

- **Data source**: National Center for Education Statistics [https://nces.ed.gov/datalab/powerstats/121-national-postsecondary-student-aid-study-2016-undergraduates/averages-medians-percents]()
- **Data description**: - this data table includes the "cumulative amount borrowed for undergrad (average) by Race/ethnicity and gender". This data was collected in the 2016 addition of the National Postsecondary Student Aid Study.

```{r}
Cum_amt_borrowed_2016 <- read_excel("amt_borrowed.xlsx", sheet = "Sheet1")
#head(Cum_amt_borrowed_2016)

Cum_amt_borrowed_2016 %>% 
  ggplot(aes(x = fct_reorder(Race, Amount, .desc = T), y = Amount, fill = Gender)) +
  geom_bar(stat="identity", position = position_dodge()) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust=1, size = 8.5, color = "black")) +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "", y = "", fill = "", title = "Average amount borrowed for undergrad in 2016 (USD) by race and gender")
```

- **Findings**: It appears that on average, females were having to borrow more for undergrad in 2016, and so were individuals who identified as Black or African American. 


### Who recieves the most aid in paying for college?

- **Data source**: National Center for Education Statistics [https://nces.ed.gov/datalab/powerstats/121-national-postsecondary-student-aid-study-2016-undergraduates/averages-medians-percents]()
- **Data description**: this data table includes the "average total aid amount for undergrad by Race/ethnicity and gender". This data was collected in the 2016 addition of the National Postsecondary Student Aid Study.

```{r}
total_aid_2016 <- read_excel("total_aid.xlsx")
#head(total_aid_2016)

total_aid_2016 %>% 
  ggplot(aes(x = fct_reorder(`Race/ethnicity`, `Total aid amount`, .desc = T), y = `Total aid amount`, fill = Gender)) +
  geom_bar(stat="identity", position = position_dodge()) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust=1, size = 8.5, color = "black")) +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "", y = "", fill = "", title = "Average total aid amount for undergrad in 2016 (USD) by race and gender")
```

- **Findings**: Average aid amounts are distributed pretty equally between gender and race categories, in 2016.


### How does the amount borrowed for college vary by major and parent's income?

- **Data Source**: - **Data source**: National Center for Education Statistics [https://nces.ed.gov/datalab/powerstats/121-national-postsecondary-student-aid-study-2016-undergraduates/averages-medians-percents]()
- **Data Description**: Data includes mean parent's income in USD in 2016, and the mean cumulative amount borrowed for college, by major.

```{r}
#devtools::install_github("slowkow/ggrepel")
library(ggrepel)
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
```

```{r}
vizdata <- read_csv("viz1data.csv")
```

```{r}
vizdata1 <- vizdata %>%
  filter(field !="Total")
```

```{r, fig.width=10.25, fig.height=5}
set.seed(123)
ggplot(vizdata1, aes(x=mean_borrow,y=mean_income,label=field, color=major_category)) +
  geom_point(size = 2.5) +
  geom_text_repel(aes(label=field), 
                  size=2.25, 
                  box.padding = 0.5, 
                  segment.size = 0.25,
                  color = "black") +
  xlim(3500,15000) + 
  ylim(60000,110000) +
  theme_classic() +
  labs(x = "Mean Cumulative Amount Borrowed (USD)", 
       y = "Mean Parents' Income (USD)", 
       title = "Amount Borrowed for Undergrad Majors Compared to Parents' Income, 2016",
       color = "Major Category") +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "right")
```

- **Findings**: After grouping the majors by category, we notice some clustering and interesting trends. We see a similar pattern for STEM educations, and also see a general negative relationship between the x and y variables if we ignore some of  the left-most "outliers".


### What majors tend to lead to higher paying jobs after college?

- **Data Source**: 
- **Data Description**:

```{r}
salary_by_major_2018 <- read_excel("salary_by_major_0818.xlsx")
#head(salary_by_major_2018)

salary_by_major_2018 %>% 
  filter(`Field of study: undergraduate (10 categories)` != "Undeclared") %>% 
  ggplot(aes(x = fct_reorder(`Field of study: undergraduate (10 categories)`, `Current job, as of B&B:08/18 interview: Annualized salary`, .desc = F), y = `Current job, as of B&B:08/18 interview: Annualized salary`)) +
  geom_bar(stat="identity", fill = "#69b3a2") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1, size = 7, color = "black")) +
  labs(x = "Undergraduate Field of Study, 2008", y = "Current Job Salary in 2018", title = "")
```

- **Findings**:
 

# Questions

1. Which of these questions seems most interesting and you think we should focus on / explore more?

2. What other background information about this topic should we consider? Are there other pieces of information we should look at or take into consideration in our analysis?

3. Do you have any other feedback about our project? What's going well and what could be improved?

